---
title: "Untitled"
author: "Yuxin Zhai"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r}
# remove object from my environment
rm(list = ls())

# Load necessary libraries
library(tidyverse)
library(readxl)
library(caret)
library(glmnet)

file_path <- "Desktop/Data-Covid002.csv"
data <- read.csv(file_path, stringsAsFactors = FALSE)
var_desc_path <- "Desktop/PPHA_30545_MP03-Variable_Description.xlsx" 
var_desc <- read_excel(var_desc_path)
# Select relevant variables
relevant_vars <- var_desc %>%
filter(Source %in% c("Opportunity Insights", "PM COVID")) %>% pull(Variable)
# Include county, state, and deathspc
relevant_vars <- c(relevant_vars, "county", "state", "deathspc") # Filter dataset
data_filtered <- data %>% select(all_of(relevant_vars))
# Summary statistics
summary_stats <- summary(data_filtered) 
print(summary_stats)

```
```{r}
# Drop rows with missing values
data_cleaned <- na.omit(data_filtered) 
str(data_cleaned)
```
```{r}
# Remove non-numeric columns before clustering
numeric_data <- data_cleaned %>%
  select(-c(county, state))  # Exclude categorical variables
numeric_data <- numeric_data %>%
  mutate(across(everything(), as.numeric))
# Normalize the data (scale to mean 0, variance 1)
numeric_scaled <- scale(numeric_data)

# Determine the optimal number of clusters using the Elbow Method
fviz_nbclust(numeric_scaled, kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal Clusters")
ggsave("Elbow_Method.png", width = 8, height = 6, dpi = 300)

set.seed(42)
kmeans_result <- kmeans(numeric_scaled, centers = 3, nstart = 10)
data_cleaned$cluster <- as.factor(kmeans_result$cluster)
```

```{r}
# Household Income vs. COVID-19 Deaths per Capita
plot1 <- ggplot(data_cleaned, aes(x = hhinc00, y = deathspc, color = cluster)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("red", "blue", "green")) +
  labs(title = "Economic Impact on COVID-19 Mortality", 
       x = "Household Income ($)", y = "Deaths per Capita", color = "Cluster") +
  theme_minimal()

print(plot1)  # Print plot
ggsave("Income_vs_DeathRate.png", plot = plot1, width = 8, height = 6, dpi = 300) 
```
Interpretation of Economic Impact on COVID-19 Mortality Clustering:
This scatter plot represents the relationship between household income and COVID-19 deaths per capita, with data points grouped into three clusters using K-Means clustering. The X-axis represents household income, while the Y-axis represents COVID-19 deaths per capita. The different colors denote clusters:
Cluster 1 (Red): Counties with high mortality rates, widely distributed across different income levels.
Cluster 2 (Blue): Counties with moderate mortality rates, concentrated mostly in lower-to-middle-income ranges.
Cluster 3 (Green): Counties with lower mortality rates, primarily located in middle-income groups.
Key Insights:
1. Higher-income does not always mean lower mortality: Unlike the previous plot, Cluster 1 (Red, high mortality) includes many counties in the higher-income bracket ($40,000 - $70,000). This suggests that factors beyond income, such as healthcare access, comorbidities, and regional pandemic policies, are influencing mortality rates.
2. Lower-income counties still show significant mortality risks: Cluster 2 (Blue) contains many low-to-middle-income counties ($15,000 - $40,000) with moderate death rates, reinforcing that economic disadvantage contributes to COVID-19 severity but is not the sole determinant.
3. Counties in Cluster 3 (Green, low mortality) are mostly concentrated at the lower range of deaths per capita: These counties are spread across different income levels but tend to have better overall health outcomes, possibly due to lower population density, stronger healthcare infrastructure, or better public health policies.
Policy Recommendations Based on K-Means Clustering:
1.Target high-mortality counties (Cluster 1) for urgent interventions, regardless of income level. These counties need enhanced healthcare infrastructure, expanded vaccine access, and better preparedness for future pandemics.
2. Investigate additional risk factors influencing high-income counties with high mortality. Since some wealthier counties still suffer high death rates, factors such as prevalence of pre-existing conditions (obesity, diabetes), urbanization, and healthcare system efficiency should be further analyzed.
3. Provide financial and healthcare support for low-income counties in Cluster 2 (Blue). These counties are at a moderate risk and could benefit from economic assistance, public health awareness programs, and increased medical resource allocation.

```{r}
unique(data_cleaned$bmi_obese_q1)
sum(data_cleaned$bmi_obese_q1 == 0)
data_cleaned %>%
  filter(bmi_obese_q1 == 0) %>%
  select(county, state, bmi_obese_q1, deathspc) %>%
  print()
sum(is.na(data$bmi_obese_q1))

selected_features <- c("bmi_obese_q1", "deathspc")
df_cluster <- data_cleaned %>%
  select(all_of(selected_features)) %>%
  na.omit()
df_cluster$cluster <- as.factor(kmeans_result$cluster)
df_cluster$cluster <- as.factor(kmeans_result$cluster)
plot_obesity <- ggplot(df_cluster, aes(x = bmi_obese_q1, y = deathspc, color = cluster)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("red", "blue", "green")) +
  labs(title = "Obesity Rate vs. COVID-19 Deaths", 
       x = "Obesity Rate", y = "Deaths per Capita", color = "Cluster") +
  theme_minimal()

print(plot_obesity)
ggsave("Obesity_vs_DeathRate.png", plot = plot_obesity, width = 8, height = 6, dpi = 300)
```
Interpretation of Obesity Rate vs. COVID-19 Mortality Clustering:
This scatter plot visualizes the relationship between obesity rates and COVID-19 deaths per capita, with counties grouped into three clusters using K-Means clustering. The X-axis represents obesity rates, while the Y-axis represents COVID-19 deaths per capita. The different colors represent the clusters:
Cluster 1 (Red): Counties with higher COVID-19 mortality, primarily falling within moderate obesity rates (~0.10 - 0.50).
Cluster 2 (Blue): Counties with low mortality but concentrated at an obesity rate of 0.00. This suggests these counties either have missing or inaccurately reported obesity data.
Cluster 3 (Green): Counties with lower mortality rates, spread across various obesity levels, including some with high obesity rates.

Key Insights:
1. Obesity is correlated with higher COVID-19 mortality, but not exclusively. Cluster 1 (Red, high-mortality counties) consists mostly of counties with moderate obesity rates (0.10 - 0.50). This supports medical research that obesity increases the risk of severe COVID-19 outcomes, but other factors like healthcare access, socio-economic conditions, and chronic illnesses may also be influencing mortality.
2. Cluster 2 (Blue) counties with 0.00 obesity rates still exist but with low mortality. Despite replacing 0.00 values with NA and re-running the clustering, these counties still appear as a distinct group. This suggests that these counties may not have reported obesity data correctly or have exceptionally low obesity rates.
3. Counties with high obesity rates (~0.50+) are mostly in Cluster 3 (Green) with lower mortality. Some high-obesity counties do not have high COVID-19 deaths, which could indicate better healthcare systems, stronger public health policies, or lower population density as protective factors.
4. Some low-obesity counties still experience high mortality. A few Cluster 1 (Red) counties with obesity rates under 0.25 still have high deaths per capita. This suggests that other factors, such as diabetes prevalence, air pollution (PM2.5), or healthcare disparities, may contribute to high COVID-19 mortality.
